---
title: "Cross-Tabulations and Grouped Bar Plots"
format: 
  live-html: 
    code-fold: true
    code-summary: "Show code"
engine: knitr
---

```{webr}
#| context: setup
#| autorun: false
#| message: false
#| warning: false
download.file(
  "https://raw.githubusercontent.com/datasciencedojo/datasets/master/titanic.csv",
  "titanic.csv"
)
library(tidyverse)
library(ggplot2)
df <- read_csv("titanic.csv")
```


# Visualizing two variables
```{r}
#| include: false
download.file(
  "https://raw.githubusercontent.com/datasciencedojo/datasets/master/titanic.csv",
  "titanic.csv"
)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(gridExtra)
df <- read_csv("titanic.csv")
```

```{r}
#| fig-width: 10
#| fig-height: 5
surv_cols <- c("Died" = "slategrey", "Survived" = "mediumseagreen")
cls_cols  <- c("1" = "#A2005699", "2" = "#5F559B99", "3" = "#00828099")

bg <- "gray95"

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      axis.title = element_text(size = 14),
      axis.text  = element_text(size = 12),
      plot.title = element_text(size = 15, face = "bold"),
      plot.background  = element_rect(fill = bg, color = NA),
      panel.background = element_rect(fill = bg, color = NA),
      plot.margin = margin(5, 10, 5, 0),
      legend.position = "none"
    )
)

tbl_theme <- ttheme_minimal(
  base_size = 12,
  core    = list(bg_params = list(fill = bg, col = NA)),
  colhead = list(bg_params = list(fill = bg, col = NA),
                 fg_params = list(fontface = "bold"))
)

tg <- function(d) {
  wrap_elements(full = tableGrob(d, rows = NULL, theme = tbl_theme)) +
    theme(plot.margin = margin(5, 0, 5, 5))
}

surv <- df %>%
  mutate(Survived = if_else(Survived == 1, "Survived", "Died")) %>%
  count(Survived, name = "Count") %>%
  mutate(Proportion = round(Count / sum(Count), 3))

cls <- df %>%
  count(Pclass, name = "Count") %>%
  mutate(Proportion = round(Count / sum(Count), 3))

p1 <- ggplot(surv, aes(Survived, Count, fill = Survived)) +
  geom_col() +
  scale_fill_manual(values = surv_cols) +
  labs(title = "Survival", x = NULL, y = "Passengers")

p2 <- ggplot(cls, aes(factor(Pclass), Count, fill = factor(Pclass))) +
  geom_col() +
  scale_fill_manual(values = cls_cols) +
  labs(title = "Passenger class", x = "Class", y = "Passengers")

(tg(surv) | p1) / (tg(cls) | p2) +
  plot_layout(widths = c(0.8, 2)) +
  plot_annotation(
    title = "Titanic Passenger Demographics",
    theme = theme(
      plot.title = element_text(size = 20, face = "bold", hjust = 0.5,
                                margin = margin(b = 12)),
      plot.background = element_rect(fill = bg, color = NA),
      plot.margin = margin(10, 10, 10, 10)
    )
  )
```

So far, we've analyzed the distribution of passengers' survival status and their class, separately. However, seeing this data, we may raise questions about the intersection of these variables -- were first-class passengers more likely to survive? How much of a difference did class make in survival rate? How much more likely would a third-class passenger have survived if they were second-class? 



## Cross-tabulations
Cross-tabulations can help us answer these questions. Cross-tabulations are essentially frequency tables, but for two variables instead of one. 
You can generate a cross-tabluation with the table() function. The inputs are the column nmaes of the two variables you want to compare. 

```{webr}
#| autorun: false
table(df$Survived, df$Pclass) 
```

This breaks down our dataset into six groups: those in first class and survived, those in first class and did not survive, those in second class and survived, and so on. 

### Improving readability 

As always, the base output is not very easy to parse. We can relabel our table using our familiar functions. 

```{webr}
#| autorun: false
xtab <- df %>%
  mutate(
    Survived = if_else(Survived == 1, "Survived", "Died"),
    Pclass   = case_when(
      Pclass == 1 ~ "First",
      Pclass == 2 ~ "Second",
      Pclass == 3 ~ "Third"
    )
  ) %>%
  count(Pclass, Survived, name = "Count")
```

## Stacked Bar Plots 

We can show this same result with a stacked bar plot.
Note that in the below example, `Pclass` is assigned to x, and `Survived` is assigned to fill. 
This means that the x-axis, or, what each bar represents, will be a passenger class. The colors will indicate the distribution of survivorship. 

```{webr}
ggplot(df, aes(x = Pclass, fill = Survived)) +
  geom_bar(position = "stack")

```

We can make a plot where the bars are survival status instead, and they are broken down by passenger class, by assigning `Survived` to x, and `Pclass` to fill. 

```{webr}
ggplot(df, aes(x = Survived, fill = Pclass)) +
  geom_bar(position = "stack")

```

### Improving readability 

```{webr}
ggplot(df, aes(x = factor(Pclass), fill = factor(Survived))) +
  geom_bar(position = "stack") +
  scale_fill_manual(
    values = c("0" = "lightblue1", "1" = "lightgray"),
    labels = c("0" = "Died", "1" = "Survived")
  ) +
  labs(
    title = "Survival by passenger class",
    x = "Passenger class",
    y = "Number of passengers",
    fill = "Outcome"
  )
```