---
title: "Numeric, Logical, Character"
format: live-html
engine: knitr
---
{{< include /_extensions/r-wasm/live/_knitr.qmd >}}
## Numeric
A numeric is a number R can do math with. 
```{webr}
#| autorun: false
60 + 7
```


### Numeric Operations

```{r}
#| echo: false
library(gt)

ops <- data.frame(
  Operator  = c("+", "-", "*", "/", "^"),
  Operation = c("addition", "subtraction", "multiplication",
                "division", "exponentiation"),
  Example   = c("5 + 2", "5 - 2", "5 * 2", "5 / 2", "5 ^ 2")
)
ops$Result <- sapply(ops$Example, \(x) eval(parse(text = x)))

gt(ops) |>
  tab_header(title = "Arithmetic operators in R") |>
  cols_align("center", columns = c(Operator, Example, Result)) |>
  tab_style(
    style = cell_text(font = "monospace"),
    locations = cells_body(columns = c(Operator, Example, Result))
  ) |>
  opt_stylize(style = 6)
```


## Logical
A logical is a TRUE or FALSE value


```{webr}
#| autorun: false
1 < 2
```


### Logical Operations

```{r}
#| echo: false
library(gt)

logic_ops <- data.frame(
  Operator  = c("<", ">", "<=", ">=", "==", "!=", "!x", "x | y", "x & y"),
  Operation = c("less than", "greater than", "less than or equal to",
                "greater than or equal to", "equal to", "not equal to",
                "not x", "x or y", "x and y"),
  Example   = c("5 < 3", "5 > 3", "5 <= 5", "5 >= 6", "5 == 5", "5 != 5",
                "!TRUE", "TRUE | FALSE", "TRUE & FALSE")
)
logic_ops$Result <- sapply(logic_ops$Example, \(x) eval(parse(text = x)))

gt(logic_ops) |>
  tab_header(title = "Comparison and logical operators in R") |>
  cols_align("center", columns = c(Operator, Example, Result)) |>
  tab_style(
    style = cell_text(font = "monospace"),
    locations = cells_body(columns = c(Operator, Example, Result))
  ) |>
  opt_stylize(style = 6)
```

## Character 
A character is text wrapped in quotes, like "hello" or "5". R treats it as words, not numbers, so you can combine, search, or change it, but you can't do math with it.


```{webr}
#| autorun: false
print("hello")
print("5")

```


# Functions 
A function is a command that takes inputs (called arguments) inside the parentheses, does something with them, and gives you back a result. This is how R knows what to do with print("hello"). 

Without the parantheses, your computer doesn't know what you're referencing.


```{webr}
#| autorun: false

print hello 
print "hello"
print() hello 
```


## The class() function 

The class function allows us to check what class the input (what we want in the parantheses) is. 


```{webr}
#| autorun: false

class(1)
class(TRUE)
class("so true")

```


It can help with diagnosing errors.


```{webr}
#| autorun: false

"12*3"
class("12*3")
```


# Assigning variables 
Like in algebra, we can assign values to variables, e.g. we set `x` equal to 5 so that `x` + 1 = 6. 
In R, we assign with `<-`

```{webr}
#| autorun: false

x <- 5 
print (x)
class (x)

print (x + 1) 
print (x<=42)
```


## You have to declare the value of a variable before doing anything with it. 


```{webr}
#| autorun: false

print (y)

y = "I'm just some characters..."
```


## Overwriting variables 

If you define a variable multiple times, it'll use the vavlue it was last defined on. 


```{webr}
#| autorun: false

z <- 1
print (z + 2)
z <- 4
print (z + 2)
```
